draft

### 2025-05-20 draft.tex
docs: Update validation text
Add appendix

% Created 2025-05-20 Tue 11:00
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features \citep{waltScikitimageImageProcessing2014}. Tools like ScaleFex or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} can generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The latest deep learning models can streamline image processing. Nevertheless, their capacity to provide biological interpretation is limited \citep{moenDeepLearningCellular2019}. In some cases, engineered features can even outperform learned ones \citep{tangMorphologicalProfilingDrug2024}. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unless a major change in neural networks occurs, they will still require more complex interpretability methods. Unlike learned features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

We first performed the numerical validation of cp\_measure relative to the original CellProfiler features. For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.

The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols. This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\end{document}

>>>>

% Created 2025-05-20 Tue 11:40
% Intended LaTeX compiler: pdflatex
\documentclass{article}

@@

Object featurization is not exclusive of CellProfiler. For example, scikit-image can produce some overlapping features \citep{waltScikitimageImageProcessing2014}. Tools like ScaleFex or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} can generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices. We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The latest deep learning models can streamline image processing. Nevertheless, their capacity to provide biological interpretation is limited \citep{moenDeepLearningCellular2019}. In some cases, engineered features can even outperform learned ones \citep{tangMorphologicalProfilingDrug2024}. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unless a major change in neural networks occurs, they will still require more complex interpretability methods. Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

We first performed the numerical validation that cp\_measure features, for a set of images and masks representative of the possible values, match the original CellProfiler features. For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each. To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features. Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler. Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.

The initial validation of our cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}. The left panel shows examples of the comparison of our features against CellProfiler's. The straight lines demonstrate the recapitulation of measurements from our implementation. A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool. The panel on the right shows the \(R^2\) value of a linear interpolation. Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols. This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
\subsection{Results}
\label{sec:orge5b5c6b}
\subsubsection{Astrocytes 3D data}
\label{sec:org447090b}

@@

\bibliographystyle{icml2025}
\bibliography{bibliography}
\section{Appendix}
\label{sec:orgdd18dd8}
\subsection{Methods}
\label{sec:orgb3e9382}
\subsubsection{Data and software}
\label{sec:orgbda0ae2}
The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
\end{document}
